---
title: "Data Exploration"
author: "Monisha Eadlaa"
date: "4/19/2020"
output:
  word_document: default
  pdf_document: default
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Set your working directory
setwd("~/Eadala_Project_Environmental_Data_Analytics")
# Load your packages
library(tidyverse)
library(knitr)
# Set your ggplot theme
mytheme <- theme_light(base_size = 12) +
  theme(plot.background = element_rect(fill = "grey97"),
        panel.grid.major =element_line(linetype = "dotted"),
        panel.grid.minor = element_line(linetype = "dotted"), text=element_text(size = 14, 
color = "black", face = "bold"),
        axis.text = element_text(color = "grey40"), 
        legend.position = "right",
        legend.text = element_text(color = "grey40"))
theme_set(mytheme)

```

# Uploading the Nutrients and ChemistryPhysics raw data files associated with the NTL-LTER lakes datasets and formating the date column of each.

```{r}
# Load your datasets
Nutrients_raw <- read.csv("./Data/Raw/NTL-LTER_Lake_Nutrients_Raw.csv")
ChemistryPhysics_raw <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

head(Nutrients_raw)
Nutrients_raw$sampledate <- as.Date(Nutrients_raw$sampledate,
format = "%m/%d/%y")

head(ChemistryPhysics_raw)
ChemistryPhysics_raw$sampledate <- as.Date(ChemistryPhysics_raw$sampledate,
format = "%m/%d/%y")

```

# Data exploration of the Nutrients and ChemistryPhysics raw data files associated with the NTL-LTER lakes datasets

```{r}
dim(Nutrients_raw)
dim(ChemistryPhysics_raw)
str(Nutrients_raw)
str(ChemistryPhysics_raw)
colnames(Nutrients_raw)
colnames(ChemistryPhysics_raw)
summary(Nutrients_raw)
summary(ChemistryPhysics_raw)

```

## Plots


```{r}
TN.FreqPol.plot <- ggplot(Nutrients_raw) +
  geom_freqpoly(aes(x = tn_ug, color = lakename), bins = 50) +
  xlab("Total Nitrogen Concentration (µg/l)") + # To label the x axis
  ylab("Count") + # To label the y axis
  labs (color = "Lake Name") # To label the legend
print(TN.FreqPol.plot)

TN.Box.plot <- ggplot(Nutrients_raw) +
  geom_boxplot(aes(x = lakename, y = tn_ug, color = lakename)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  theme(legend.position="none") + # To removes the legend
  xlab("Lake Name") + # To label the x axis
  ylab("TN (µg/l)") # To label the y axis
print(TN.Box.plot)

TN.Scatter.plot <- ggplot(Nutrients_raw, aes(x = sampledate, y = tn_ug, color = lakename)) +
  geom_point() +
  stat_smooth(color = "black") +
  xlab("Lake Name") + # To label the x axis
  ylab("TN (µg/l)") + # To label the y axis
  labs (color = "Lake Name") # To label the legend
print(TN.Scatter.plot)

DO.Scatter.plot <- ggplot(ChemistryPhysics_raw, aes(x = sampledate, y = dissolvedOxygen, color = lakename)) +
  geom_point() +
  stat_smooth(color = "black") +
  ylim(0,25) + # To remove the outliers
  xlab("Lake Name") + # To label the x axis
  ylab("Dissolved Oxygen (mg/L)") + # To label the y axis
  labs (color = "Lake Name") # To label the legend
print(DO.Scatter.plot)
```

# Data Exploring: What is the average nitrogen:phosphorus ratio for each of the six north temperate lakes? Are the lakes phosphorus or nitrogen deficient? 

A Redfield ratio N:P of 16:1 by moles in general indicates a roughly balanced supply of N and P, and algae assemblages tend to mirror this ratio fairly closely when growing under balanced growth conditions (https://www.sciencedirect.com/topics/earth-and-planetary-sciences/redfield-ratio). Converting the molar ratio to migrogram ratio gives us N:P of 7.24:1, since 1 mole of N = 14.0067g and 1 mole of P = 30.973761g (https://www.convertunits.com/from/grams+Nitrogen/to/moles; https://www.convertunits.com/from/grams+Phosphorus/to/moles)

```{r}
#To exlude some of the valariables from the dataset so that the number of NAs in the dataset could be minimized
NitrogenPhosphorus <- Combined_processed %>%
  select(-temperature_C) %>%
  na.exclude()

# Save the new folder
write.csv(NitrogenPhosphorus, row.names = FALSE, file = "./Data/Processed/NitrogenPhosphorus_processed.csv")

# To understand if most of the lakes are above or below the Redfield Ratio - to understand of most of the lakes are phosphorus or nitrogen limited
NP.plot <- ggplot(NitrogenPhosphorus, aes(x = tp_ug, y = tn_ug, color = lakename)) +
  geom_point(alpha = 0.75) +
  geom_abline(slope = 7.24, intercept = 0, linetype = "dashed") +
  annotate("text", x = 40, y = 30 * 7.24, angle = atan(7.24 * 4) * 35/pi, label = "Redfield Ratio", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_color_brewer(palette = "Set1", direction = 1) +
  ggtitle("(a)") + # To insert the main title
  theme(plot.title = element_text(size = 14, face = "bold")) +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=12,face="bold")) + 
  theme(legend.title = element_text(size = 12), legend.text=element_text(size=12)) +
  xlab("P (µg/l)") + # To label the x axis
  ylab("N (µg/l)") + # To label the y axis
  labs (color = "Lake Name") + # To label the legend
  xlim(0,65)
print(NP.plot) # Gives the N vs P plot #Tells us that most of the Lakes are phosphorus deficient, therefore phosphorus should be the limiting nutrient addressed in the water management

NitrogenPhosphorus$NPRatio <- NitrogenPhosphorus$tn_ug / NitrogenPhosphorus$tp_ug #gives us a new colum with the N:P ratio

# To summarize the N:P ratio in each lake 
NPRatio.summary <- NitrogenPhosphorus %>%
   group_by(lakename) %>%
  summarise(mean.NPRatio = mean(NPRatio, na.rm = TRUE),
            minimum.NPRatio = min(NPRatio, na.rm = TRUE),
            maximum.NPRatio = max(NPRatio, na.rm = TRUE),
            Standard.dev.NPRatio = sd(NPRatio, na.rm = TRUE)) # Gives us summary stats of each of the lakes with respect to N:P

# to understand the distriution of N:P ratio in each lake
NPRatio.boxplot <- ggplot(NitrogenPhosphorus, aes(x = depth, y = NPRatio)) +
  geom_boxplot(aes(color = lakename)) +
  scale_color_brewer(palette = "Set1", direction = 1) +
  geom_hline(yintercept = 7.24, linetype = "dashed") +
  annotate("text", x=4, y=2, label="Redfield Ratio", size = 3) +
  ggtitle("(b)") + # To insert the main title 
  theme(plot.title = element_text(size = 14, face = "bold")) +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=12,face="bold")) + 
  theme(legend.title = element_text(size = 12), legend.text=element_text(size=12)) +
  xlab("Depth (m)") + # To label the x axis
  ylab("N:P") + # To label the y axis
  labs (color = "Lake Name") + # To label the legend
  ylim(0,200)
print(NPRatio.boxplot) # Gives us the individual boxplots of the lakes with respect to their N:P

# Since all the lakes are phosphoshorus deficient, has water management practices managed to reduce the phosphorus over the years?
NP.years.plot <- ggplot(NitrogenPhosphorus, aes(x = year4, y = tp_ug, color = lakename)) +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_color_brewer(palette = "Set1", direction = 1) +
  ggtitle("") + # To insert the main title
  theme(legend.title = element_text(size = 12), legend.text=element_text(size=12)) +
  xlab("Years") + # To label the x axis
  ylab("TP (µg/l)") + # To label the y axis
  labs (color = "Lake Name") # To label the legend
print(NP.years.plot) # Gives the N vs P plot; tells us that there has been a decrease in phosphorus concentration over the years.
```

Figure for slack showcase.

```{r fig1, echo=FALSE,message=FALSE,include=TRUE,fig.cap=" (a) Total nitrogen concentration versus total phosphorus concentration from the 6 Wisconsin lakes data collected over the years 1991–2016. Dashed line indicates the Redfield Ratio of 7.24:1, which indicates optimal conditions for phytoplankton growth. (b) Distribution of N:P ratios in the same dataset."}
library(cowplot)

NP.plot = NP.plot + theme(legend.position="none") # Removes the legend from the plot
NPRatio.boxplot = NPRatio.boxplot + theme(legend.position ="right", legend.title = element_text(size = 7), legend.text = element_text(size = 7)) # Asserts the position of the legend on the right side of the plot
plot_grid(NP.plot, NPRatio.boxplot, nrow = 1, align = 'h', rel_widths = c(1, 1.45)) # Combines all the three plots

```

