---
title: "Data Analysis"
author: "Monisha Eadala"
date: "4/19/2020"
output:
  pdf_document: default
  word_document: default
  html_document: default
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
setwd("~/Eadala_Project_Environmental_Data_Analytics")
# Load your packages
library(tidyverse)
library(knitr)
# Set your ggplot theme
mytheme <- theme_light(base_size = 12) +
  theme(plot.background = element_rect(fill = "grey97"),
        panel.grid.major =element_line(linetype = "dotted"),
        panel.grid.minor = element_line(linetype = "dotted"), text=element_text(size = 14, 
color = "black", face = "bold"),
        axis.text = element_text(color = "grey40"), 
        legend.position = "right",
        legend.text = element_text(color = "grey40"))
theme_set(mytheme)

Nutrients_raw <- read.csv("./Data/Raw/NTL-LTER_Lake_Nutrients_Raw.csv")
ChemistryPhysics_raw <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

head(Nutrients_raw)
Nutrients_raw$sampledate <- as.Date(Nutrients_raw$sampledate,
format = "%m/%d/%y")

head(ChemistryPhysics_raw)
ChemistryPhysics_raw$sampledate <- as.Date(ChemistryPhysics_raw$sampledate,
format = "%m/%d/%y")

list(unique(ChemistryPhysics_raw$year4))
list(unique(Nutrients_raw$year4))

Nutrients_processed <- Nutrients_raw %>%
  select(-lakeid, -depth_id , -nh34, -no23, -po4, -comments) %>%
  filter(year4 == 1991 | year4 == 1992 | year4 ==  1993 | year4 ==  1994 | year4 == 1999 | year4 ==  2000 | year4 == 2001 | year4 == 2002 | year4 == 2003 | year4 == 2005 | year4 ==  2006 | year4 == 2007| year4 == 2008 | year4 == 2009 | year4 == 2010 | year4 ==  2011 | year4 ==  2012 | year4 ==  2013 | year4 ==  2014 | year4 == 2015 | year4 == 2016)

list(unique(Nutrients_processed$lakename))

ChemistryPhysics_processed <- ChemistryPhysics_raw %>%
  select(-lakeid, -irradianceDeck, -irradianceWater, -comments) %>% 
  filter(year4 == 1991 | year4 == 1992 | year4 ==  1993 | year4 ==  1994 | year4 == 1999 | year4 ==  2000 | year4 == 2001 | year4 == 2002 | year4 == 2003 | year4 == 2005 | year4 ==  2006 | year4 == 2007| year4 == 2008 | year4 == 2009 | year4 == 2010 | year4 ==  2011 | year4 ==  2012 | year4 ==  2013 | year4 ==  2014 | year4 == 2015 | year4 == 2016)
list(unique(ChemistryPhysics_processed$lakename))

Combined_processed <- full_join(Nutrients_processed, ChemistryPhysics_processed) %>%
  filter(year4 == 1991 | year4 == 1992 | year4 ==  1993 | year4 ==  1994 | year4 == 1999 | year4 ==  2000 | year4 == 2001 | year4 == 2002 | year4 == 2003 | year4 == 2005 | year4 ==  2006 | year4 == 2007| year4 == 2008 | year4 == 2009 | year4 == 2010 | year4 ==  2011 | year4 ==  2012 | year4 ==  2013 | year4 ==  2014 | year4 == 2015 | year4 == 2016) %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake" | lakename == "East Long Lake" | lakename == "West Long Lake" | lakename == "Tuesday Lake" | lakename == "Crampton Lake")

write.csv(Combined_processed, row.names = FALSE, file = "./Data/Processed/Combined_processed.csv")
#To exlude some of the valariables from the dataset so that the number of NAs in the dataset could be minimized
NitrogenPhosphorus <- Combined_processed %>%
  select(-temperature_C) %>%
  na.exclude()

# To understand if most of the lakes are above or below the Redfield Ratio - to understand of most of the lakes are phosphorus or nitrogen limited
NP.plot <- ggplot(NitrogenPhosphorus, aes(x = tp_ug, y = tn_ug, color = lakename)) +
  geom_point(alpha = 0.75) +
  geom_abline(slope = 7.24, intercept = 0, linetype = "dashed") +
  annotate("text", x = 40, y = 30 * 7.24, angle = atan(7.24 * 4) * 35/pi, label = "Redfield Ratio", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_color_brewer(palette = "Set1", direction = 1) +
  ggtitle("(a)") + # To insert the main title
  theme(plot.title = element_text(size = 14, face = "bold")) +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=12,face="bold")) + 
  theme(legend.title = element_text(size = 12), legend.text=element_text(size=12)) +
  xlab("P (µg/l)") + # To label the x axis
  ylab("N (µg/l)") + # To label the y axis
  labs (color = "Lake Name") + # To label the legend
  xlim(0,65)
print(NP.plot) # Gives the N vs P plot #Tells us that most of the Lakes are phosphorus deficient, therefore phosphorus should be the limiting nutrient addressed in the water management

NitrogenPhosphorus$NPRatio <- NitrogenPhosphorus$tn_ug / NitrogenPhosphorus$tp_ug #gives us a new colum with the N:P ratio

# To summarize the N:P ratio in each lake 
NPRatio.summary <- NitrogenPhosphorus %>%
   group_by(lakename) %>%
  summarise(mean.NPRatio = mean(NPRatio, na.rm = TRUE),
            minimum.NPRatio = min(NPRatio, na.rm = TRUE),
            maximum.NPRatio = max(NPRatio, na.rm = TRUE),
            Standard.dev.NPRatio = sd(NPRatio, na.rm = TRUE)) # Gives us summary stats of each of the lakes with respect to N:P

# to understand the distriution of N:P ratio in each lake
NPRatio.boxplot <- ggplot(NitrogenPhosphorus, aes(x = depth, y = NPRatio)) +
  geom_boxplot(aes(color = lakename)) +
  scale_color_brewer(palette = "Set1", direction = 1) +
  geom_hline(yintercept = 7.24, linetype = "dashed") +
  annotate("text", x=4, y=2, label="Redfield Ratio", size = 3) +
  ggtitle("(b)") + # To insert the main title 
  theme(plot.title = element_text(size = 14, face = "bold")) +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=12,face="bold")) + 
  theme(legend.title = element_text(size = 12), legend.text=element_text(size=12)) +
  xlab("Depth (m)") + # To label the x axis
  ylab("N:P") + # To label the y axis
  labs (color = "Lake Name") + # To label the legend
  ylim(0,200)
print(NPRatio.boxplot) # Gives us the individual boxplots of the lakes with respect to their N:P

# Since all the lakes are phosphoshorus deficient, has water management practices managed to reduce the phosphorus over the years?
NP.years.plot <- ggplot(NitrogenPhosphorus, aes(x = year4, y = tp_ug, color = lakename)) +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_color_brewer(palette = "Set1", direction = 1) +
  ggtitle("") + # To insert the main title
  theme(legend.title = element_text(size = 12), legend.text=element_text(size=12)) +
  xlab("Years") + # To label the x axis
  ylab("TP (µg/l)") + # To label the y axis
  labs (color = "Lake Name") # To label the legend
print(NP.years.plot) # Gives the N vs P plot; tells us that there has been a decrease in phosphorus concentration over the years.
```

# Data Analysis

## One-sample t-test

### Research question 1: Have the lakes been within the limiting nutrient criteria over the years? How does the N:P ratio impact the dissolved oxygen in the lakes?

The above figures tells us that all the eight lakes are phosphorus deficient, and therefore phosphorus is the element that needs to be controlled on priority to avoid eutrophication. EPA has compiled state, territorial, and authorized tribal water quality standards that EPA has approved or are otherwise in effect for Clean Water Act purposes. According to this compilation, Wisconsin's stratified lakes' total phosphorus criterion is not more than 30µg/l (https://www.epa.gov/wqs-tech/state-specific-water-quality-standards-effective-under-clean-water-act-cwa#tb2). According to Carlson R.E. and J. Simpson in 1996, phosphorus concentration between 24µg/l and 96µg/l suggests eutrophic conditions, and anything above 96µg/l suggests hypereutrophic conditions (https://en.wikipedia.org/wiki/Trophic_state_index). Therefore, it is important to check if the eight North Temperate Lakes are at least within the 30µg/l criterion or not. 

The last statistical analysis will test the null hypothesis that the means of the total phosphorus concentrations in the eight North Temperate Lakes are below the regulatory standard of 30µg/l. 

First, the assumption of normal distribution is evaluated through the Shapiro-Wilk normality test.

```{r}
shapiro.test(NitrogenPhosphorus$tp_ug) 
```

The Shapiro-Wilk normality test says that the total phoshorus concentrations data in the six lakes of Wisconsin are significantly different from a normal distribution (Shapiro-Wilk normality test; W = 0.69337, p-value < 0.0001)

Next, a visual analysis of the data is performed.

```{r, echo=FALSE}
hist(NitrogenPhosphorus$tp_ug, breaks = 50) # Phosphorus concentration histogram

qqnorm(NitrogenPhosphorus$tp_ug, pch = 1, frame = FALSE)
qqline(NitrogenPhosphorus$tp_ug, col = "blue", lwd = 2)
```

In the above figures it can be seen that the data distributuon is right-skewed with a longer tail to the right than a normal distribution; nevertheless, environmental data often violate the assumptions of normality and since the sample size is large, a t-test is performed anyway.

```{r}
t.test(NitrogenPhosphorus$tp_ug, mu = 30, alternative = "less")
```

According to the One Sample t-test, the TP concentrations for the North Temperate Lakes of Wisconsin from 1991 to 2016 have not been significantly lower than 30µg/l (one sample t-test; t = 3.2053, df = 907, p-value > 0.0001).

```{r}
TPvsCriterion <- ggplot(NitrogenPhosphorus, aes(x = year4, y = tp_ug, color = lakename)) +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_hline(yintercept = 30, linetype = "dashed") +
  scale_color_brewer(palette = "Set1", direction = 1)
print(TPvsCriterion) # This presens a visualization of the TP concentrations in Wisconsin lakes data in comparison with the Wisconsin regulatory standard.
```

The above figure presents a visualization of the TP concentrations in Wisconsin lakes data in comparison with the Wisconsin regulatory standard. The blue line in comparison with the black dashed line indicates that most of the Wisconsin lakes started off with phosphorus levels above the criteria but have been able to lower their phosphorus more recently.

Coming the the next part of the research question (how does the N:P ratio affect the dissolved oxygen?), below gives us the visual interpretation: 
```{r}
NPRatio.plot <- ggplot(NitrogenPhosphorus, aes(x = NPRatio, y = dissolvedOxygen, color = lakename)) +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_color_brewer(palette = "Set1", direction = 1) +
  ylim(0,15) +
  xlim(0,150)
print(NPRatio.plot)
```

This visual tells us that an increase in the N:P ratio increased dissolved oxygen in the lakes. This is consistent with the theory that controlling phosphorus concentration in phosphorus deficient lakes can can preserve their water quality. However, phosphorus might not be the only factor affecting the dissolved oxygen and eventutally the water quality of water. It is important to understand what other variables affect dissolved oxygen, whose depletion indicates eutrophication in most cases.

## GLM

### Research question 2: What are the best set of predictors of dissolved oxygen across the monitoring period at the North Temperate Lakes LTER
```{r}
DO.predictors <- Combined_processed %>%
  na.exclude()

DO.AIC <- lm(data = DO.predictors, dissolvedOxygen ~ year4 + daynum + temperature_C + tn_ug + tp_ug + depth) # Runs an AIC to determine what set of explanatory variables is best suited to predict dissolved oxygen
step(DO.AIC) # Confirms that I should be using the variables that correspnd to the model with the lowest AIC; Which means the best predictors are year, day, month, temperature_C, tn_ug, tp_ug and depth

DO.model <- lm(data = DO.predictors, dissolvedOxygen ~ year4 + daynum + temperature_C + depth + tn_ug + tp_ug) 
anova(DO.model) # Reconfirmed the above variables as the best

DO.model <- lm(data = DO.predictors, dissolvedOxygen ~ year4 + daynum + temperature_C + depth + tn_ug + tp_ug) # Asserts the chosen model
summary(DO.model) # Runs a multiple regression on the above recommended set of variables

Year.plot <- ggplot(DO.predictors, aes(x = year4, y = dissolvedOxygen, color = lakename)) +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_color_brewer(palette = "Set1", direction = 1) +
  xlab("Year") + # To label the x axis
  ylab("Dissolved Oxygen (mh/l)") + # To label the y axis
  labs (color = "Lake Name") # To label the legend
print(Year.plot)

Daynum.plot <- ggplot(DO.predictors, aes(x = daynum, y = dissolvedOxygen, color = lakename)) +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_color_brewer(palette = "Set1", direction = 1) +
  xlab("Day Number") + # To label the x axis
  ylab("Dissolved Oxygen (mh/l)") + # To label the y axis
  labs (color = "Lake Name") # To label the legend
print(Daynum.plot)

Temp.plot <- ggplot(DO.predictors, aes(x = temperature_C, y = dissolvedOxygen, color = lakename)) +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_color_brewer(palette = "Set1", direction = 1) +
  xlab("Temperature (C)") + # To label the x axis
  ylab("Dissolved Oxygen (mh/l)") + # To label the y axis
  labs (color = "Lake Name") + # To label the legend
  ylim(0,30)
print(Temp.plot)

Depth.plot <- ggplot(DO.predictors, aes(x = depth, y = dissolvedOxygen, color = lakename)) +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_color_brewer(palette = "Set1", direction = 1) +
  xlab("Depth (m)") + # To label the x axis
  ylab("Dissolved Oxygen (mh/l)") + # To label the y axis
  labs (color = "Lake Name") + # To label the legend
  ylim (0,25)
print(Depth.plot)

TN.plot <- ggplot(DO.predictors, aes(x = tn_ug, y = dissolvedOxygen, color = lakename)) +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_color_brewer(palette = "Set1", direction = 1) +
  xlab("N (µg/l)") + # To label the x axis
  ylab("Dissolved Oxygen (mg/l)") + # To label the y axis
  labs (color = "Lake Name") + # To label the legend
  ylim(0, 20)
print(TN.plot)

TP.plot <- ggplot(DO.predictors, aes(x = tp_ug, y = dissolvedOxygen, color = lakename)) +
  geom_point(alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  scale_color_brewer(palette = "Set1", direction = 1) +
  xlab("P (µg/l)") + # To label the x axis
  ylab("Dissolved Oxygen (mh/l)") + # To label the y axis
  labs (color = "Lake Name") + # To label the legend
  ylim(0, 20)
print(TP.plot)
```
